Channel Shaping Using Reconfigurable Intelligent Surfaces: From Diagonal to Beyond

### 2024-02-06 main.tex
:pencil: Update proposition 4

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}
\newacronym{los}{LoS}{Line-of-Sight}
\newacronym{csi}{CSI}{Channel State Information}
\newacronym{cscg}{CSCG}{Circularly Symmetric Complex Gaussian}
\newacronym{sca}{SCA}{Successive Convex Approximation}

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

		% The Pareto singular value problem \eqref{op:pareto} can thus be solved by Algorithm~\ref{ag:rcg}.
		\begin{lemma}
			The Euclidean gradient of \eqref{ob:pareto} w.r.t. \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U} \in \mathbb{C}^{N_\mathrm{R} \times N}$ and $\mathbf{V} \in \mathbb{C}^{N_\mathrm{T} \times N}$ are the left and right compact singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Let $\mathbf{H} = \sum_n \mathbf{u}_n \sigma_n \mathbf{v}_n^\mathsf{H}$ be the compact \gls{svd} of the equivalent channel.
			Since the singular vectors are orthonormal, the $n$-th singular value can be expressed as
			\begin{equation}
				\sigma_n = \mathbf{u}_n^\mathsf{H} \mathbf{H} \mathbf{v}_n = \mathbf{u}_n^\mathsf{T} \mathbf{H}^* \mathbf{v}_n^*,
			\end{equation}
			whose differential w.r.t. $\mathbf{\Theta}_g^*$ is
			\begin{align*}
				\partial \sigma_n
				& = \partial \mathbf{u}_n^\mathsf{T} \underbrace{\mathbf{H}^* \mathbf{v}_n^*}_{\sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T} \mathbf{v}_n} + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \underbrace{\mathbf{u}_n^\mathsf{T} \mathbf{H}^*}_{\mathbf{u}_n^\mathsf{T} \sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T}} \partial \mathbf{v}_n^*\\
				& = \underbrace{\partial \mathbf{u}_n^\mathsf{T} \mathbf{u}_n^*}_{\partial 1 = 0} \cdot \sigma_n + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \sigma_n \cdot \underbrace{\mathbf{v}_n^\mathsf{T} \partial \mathbf{v}_n^*}_{\partial 1 = 0}\\
				& = \mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\\
				& = \mathrm{tr}(\mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \cdot \partial \mathbf{\Theta}_g^*).
			\end{align*}
			According to \cite{Hjorungnes2007}, the corresponding complex derivative is
			\begin{equation}
				\frac{\partial \sigma_n}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{u}_n \mathbf{v}_n^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:derivative_sv}
			\end{equation}
			Linear combination of \eqref{eq:derivative_sv} yields \eqref{eq:pareto_gradient}.
		\end{proof}

@@

	\begin{subsection}{Alternating Optimization}
		Consider an \gls{ao} approach that updates $\mathbf{\Theta}$ and $\mathbf{W}$ iteratively.
		For a given $\mathbf{W}$, the passive beamforming subproblem is
		\begin{maxi!}
			% {\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{(\mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F}) \mathbf{Q} (\mathbf{H}_\mathrm{D}^\mathsf{H} + \mathbf{H}_\mathrm{F}^\mathsf{H} \mathbf{\Theta}^\mathsf{H} \mathbf{H}_\mathrm{B}^\mathsf{H})}{\eta}\biggr)}{\label{op:rate_passive}}{\label{ob:rate_passive}}
			{\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{H} \mathbf{Q} \mathbf{H}^\mathsf{H}}{\eta}\biggr)}{\label{op:rate_passive}}{\label{ob:rate_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		where $\mathbf{Q} \triangleq \mathbf{W} \mathbf{W}^\mathsf{H}$ is the transmit covariance matrix.
		\begin{lemma}
			The Euclidean gradient of \eqref{ob:rate_passive} w.r.t. \gls{bd}-\gls{ris} block $g$ is
			\begin{equation}
				\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\eta} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \biggl(\mathbf{I} + \frac{\mathbf{H}\mathbf{Q}\mathbf{H}^\mathsf{H}}{\eta}\biggr)^{-1} \mathbf{H} \mathbf{Q} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:rate_gradient}
			\end{equation}
		\end{lemma}

		\begin{proof}
			The differential of $R$ w.r.t. $\mathbf{\Theta}_g^*$ is \cite{Hjorungnes2007}
			\begin{align*}
				\partial R
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \partial \mathbf{H}^* \cdot \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \biggr\}\\
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \biggr\}\\
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \mathbf{H}_{\mathrm{F},g}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \biggr\},
			\end{align*}
			and the corresponding complex derivative is \eqref{eq:rate_gradient}.
		\end{proof}
		% Problem \eqref{op:rate_passive} is then solved by Algorithm \ref{ag:rcg} where line \ref{ln:gradient_euclidean} uses \eqref{eq:rate_gradient} explicitly.
		% Problem \eqref{op:rate_passive} is then solved by Algorithm \ref{ag:rcg} where line \ref{ln:gradient_euclidean} uses \eqref{eq:rate_gradient} explicitly.
		Algorithm \ref{ag:rcg} is then invoked to solve problem \eqref{op:rate} where line \ref{ln:gradient_euclidean} uses \eqref{eq:rate_gradient} explicitly.
		Since \eqref{ob:rate_passive} is a concave function of $\mathbf{\Theta}$, convergence to local-optimal points is guaranteed.
		On the other hand, the global optimal transmit precoder for a fixed $\mathbf{\Theta}$ is given by the eigenmode transmission \cite{Clerckx2013}
		% For a given $\mathbf{\Theta}$, the global optimal transmit precoder is given by eigenmode transmission \cite{Clerckx2013}
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
			\label{eq:precoder_eigenmode}
		\end{equation}
		where $\mathbf{V}$ is the right channel singular matrix and $\mathbf{S}^\star$ is the optimal water-filling power allocation matrix.
		The overall \gls{ao} algorithm converges to local-optimal points of problem \eqref{op:rate} since each subproblem is solved optimally and the objective is bounded from above.
		% is a first-order method to \eqref{op:rate_passive} and the water-filling algorithm is a global optimal solution to the convex problem \eqref{op:rate}.
	\end{subsection}

	\begin{subsection}{Low-Complexity Solution}
		We then propose a low-complexity solution to problem \eqref{op:rate} inspired by channel shaping.
		The passive beamforming subproblem \eqref{op:rate_passive} involves transmit covariance matrix $\mathbf{Q}$ and thus requires iterative \gls{rcg} update.
		Instead, we decouple the joint \gls{ris}-transceiver design by recasting \eqref{op:rate_passive} as channel power maximization
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which only accounts for channel characteristics but is still non-trivial.
		Inspired by \cite{Nie2017}, we successively approximate the quadratic objective \eqref{ob:power_passive} by local Taylor expansions and solve each iteration in closed form.
		% approximate the quadratic objective \eqref{ob:power_passive} with a sequence of affines and solve each problem
		% iterative algorithm to solve \eqref{op:pc_power}.
		% The idea is to successively approximate the quadratic objective with a sequence of affine functions and solve the resulting subproblems in closed form.

>>>>

% glossaries-extra
\glsdisablehyper
\setabbreviationstyle[acronym]{long-short}
\newacronym{ao}{AO}{Alternating Optimization}
\newacronym{bd}{BD}{Beyond-Diagonal}
\newacronym{bcd}{BCD}{Block Coordinate Descent}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{siso}{SISO}{Single-Input Single-Output}
\newacronym{miso}{MISO}{Multiple-Input Single-Output}
\newacronym{mimo}{MIMO}{Multiple-Input Multiple-Output}
\newacronym{rcg}{RCG}{Riemannian Conjugate Gradient}
\newacronym{ris}{RIS}{Reconfigurable Intelligent Surface}
\newacronym{pc}{PC}{Point-to-point Channel}
\newacronym{ic}{IC}{Interference Channel}
\newacronym{snr}{SNR}{Signal-to-Noise Ratio}
\newacronym{wsr}{WSR}{Weighted Sum-Rate}
\newacronym{svd}{SVD}{Singular Value Decomposition}
\newacronym{mmse}{MMSE}{Minimum Mean-Square Error}
\newacronym{wmmse}{WMMSE}{Weighted \gls{mmse}}
\newacronym{mse}{MSE}{Mean-Square Error}
\newacronym{los}{LoS}{Line-of-Sight}
\newacronym{csi}{CSI}{Channel State Information}
\newacronym{cscg}{CSCG}{Circularly Symmetric Complex Gaussian}
\newacronym{sca}{SCA}{Successive Convex Approximation}
\newacronym{kkt}{KKT}{Karush-Kuhn-Tucker}

@@

	\emph{Notation:}
	Italic, bold lower-case, and bold upper-case letters indicate scalars, vectors and matrices, respectively.
	$\jmath$ denotes the imaginary unit.
	$\mathbb{C}$ represents the set of complex numbers.
	$\mathbb{U}^{n \times n}$ denotes the set of $n \times n$ unitary matrices.
	$\mathbf{0}$ and $\mathbf{I}$ are the all-zero and identity matrices with appropriate size, respectively.
	$\Re\{\cdot\}$ takes the real part of a complex number.
	$\mathrm{tr}(\cdot)$ and $\det(\cdot)$ evaluates the trace and determinant of a square matrix, respectively.
	$\mathrm{diag}(\cdot)$ constructs a square matrix with arguments on the main diagonal and zeros elsewhere.
	$\mathrm{sv}(\cdot)$ returns the singular value vector.
	$\sigma_n(\cdot)$ and $\lambda_n(\cdot)$ is the $n$-th largest singular value and eigenvalue, respectively.
	% $\boldsymbol{\sigma}(\cdot)$ and $\boldsymbol{\lambda}(\cdot)$ are the corresponding vectors.
	$(\cdot)^*$, $(\cdot)^\mathsf{T}$, $(\cdot)^\mathsf{H}$, $(\cdot)^{(r)}$, $(\cdot)^{\star}$ denote the conjugate, transpose, conjugate transpose (Hermitian), $r$-th iterated point, and final solution, respectively.
	$(\cdot)_{[x:y]}$ is a shortcut for $(\cdot)_x,(\cdot)_{x+1},\ldots,(\cdot)_y$.
	$\lvert \cdot \rvert$ denotes the absolute value.
	% $\lVert \cdot \rVert _p$ means the $p$-norm and $\lVert \cdot \rVert$ suggests $p = 2$.
	$\lVert \cdot \rVert$ means the Euclidean norm.
	$\lVert \cdot \rVert _\mathrm{F}$ represents the Frobenius norm.
	$\mathcal{CN}(\mathbf{0}, \mathbf{\Sigma})$ is the multivariate \gls{cscg} distribution with mean $\mathbf{0}$ and covariance $\mathbf{\Sigma}$.
	$\sim$ means ``distributed as''.
\end{section}

@@

		% The Pareto singular value problem \eqref{op:pareto} can thus be solved by Algorithm~\ref{ag:rcg}.
		\begin{lemma}
			The Euclidean gradient of \eqref{ob:pareto} w.r.t. \gls{bd}-\gls{ris} group $g$ is
			\begin{equation}
				\frac{\partial \sum_n \rho_n \sigma_n(\mathbf{H})}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{U} \mathrm{diag}(\rho_1,\ldots,\rho_N) \mathbf{V}^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H},
				\label{eq:pareto_gradient}
			\end{equation}
			where $\mathbf{U} \in \mathbb{C}^{N_\mathrm{R} \times N}$ and $\mathbf{V} \in \mathbb{C}^{N_\mathrm{T} \times N}$ are the left and right compact singular matrices of $\mathbf{H}$, respectively.
		\end{lemma}
		\begin{proof}
			Let $\mathbf{H} = \sum_n \mathbf{u}_n \sigma_n \mathbf{v}_n^\mathsf{H}$ be the compact \gls{svd} of the equivalent channel.
			Since the singular vectors are orthonormal, the $n$-th singular value can be expressed as
			\begin{equation}
				\sigma_n = \mathbf{u}_n^\mathsf{H} \mathbf{H} \mathbf{v}_n = \mathbf{u}_n^\mathsf{T} \mathbf{H}^* \mathbf{v}_n^*,
			\end{equation}
			whose differential w.r.t. $\mathbf{\Theta}_g^*$ is
			\begin{align*}
				\partial \sigma_n
				& = \partial \mathbf{u}_n^\mathsf{T} \underbrace{\mathbf{H}^* \mathbf{v}_n^*}_{\sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T} \mathbf{v}_n} + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \underbrace{\mathbf{u}_n^\mathsf{T} \mathbf{H}^*}_{\mathbf{u}_n^\mathsf{T} \sum_m \mathbf{u}_m^* \sigma_m \mathbf{v}_m^\mathsf{T}} \partial \mathbf{v}_n^*\\
				& = \underbrace{\partial \mathbf{u}_n^\mathsf{T} \mathbf{u}_n^*}_{\partial 1 = 0} \cdot \sigma_n + \mathbf{u}_n^\mathsf{T} \cdot \partial \mathbf{H}^* \cdot \mathbf{v}_n^* + \sigma_n \cdot \underbrace{\mathbf{v}_n^\mathsf{T} \partial \mathbf{v}_n^*}_{\partial 1 = 0}\\
				& = \mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\\
				& = \mathrm{tr}(\mathbf{H}_{\mathrm{F},g}^* \mathbf{v}_n^*\mathbf{u}_n^\mathsf{T} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^*).
			\end{align*}
			According to \cite{Hjorungnes2007}, the corresponding complex derivative is
			\begin{equation}
				\frac{\partial \sigma_n}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \mathbf{u}_n \mathbf{v}_n^\mathsf{H} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:derivative_sv}
			\end{equation}
			Linear combination of \eqref{eq:derivative_sv} yields \eqref{eq:pareto_gradient}.
		\end{proof}

@@

	\begin{subsection}{Alternating Optimization}
		Consider an \gls{ao} approach that updates $\mathbf{\Theta}$ and $\mathbf{W}$ iteratively.
		For a given $\mathbf{W}$, the passive beamforming subproblem is
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\log \det \biggl(\mathbf{I} + \frac{\mathbf{H} \mathbf{Q} \mathbf{H}^\mathsf{H}}{\eta}\biggr)}{\label{op:rate_passive}}{\label{ob:rate_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		where $\mathbf{Q} \triangleq \mathbf{W} \mathbf{W}^\mathsf{H}$ is the transmit covariance matrix.
		\begin{lemma}
			The Euclidean gradient of \eqref{ob:rate_passive} w.r.t. \gls{bd}-\gls{ris} block $g$ is
			\begin{equation}
				\frac{\partial R}{\partial \mathbf{\Theta}_g^*} = \frac{1}{\eta} \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \biggl(\mathbf{I} + \frac{\mathbf{H}\mathbf{Q}\mathbf{H}^\mathsf{H}}{\eta}\biggr)^{-1} \mathbf{H} \mathbf{Q} \mathbf{H}_{\mathrm{F},g}^\mathsf{H}.
				\label{eq:rate_gradient}
			\end{equation}
		\end{lemma}

		\begin{proof}
			The differential of $R$ w.r.t. $\mathbf{\Theta}_g^*$ is \cite{Hjorungnes2007}
			\begin{align*}
				\partial R
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \partial \mathbf{H}^* \cdot \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \biggr\}\\
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \biggr\}\\
				& = \frac{1}{\eta} \mathrm{tr} \biggl\{ \mathbf{H}_{\mathrm{F},g}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T} \Bigl(\mathbf{I} + \frac{\mathbf{H}^* \mathbf{Q}^\mathsf{T} \mathbf{H}^\mathsf{T}}{\eta}\Bigr)^{-1} \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \biggr\},
			\end{align*}
			and the corresponding complex derivative is \eqref{eq:rate_gradient}.
		\end{proof}
		Algorithm \ref{ag:rcg} is then invoked to solve problem \eqref{op:rate} where line \ref{ln:gradient_euclidean} uses \eqref{eq:rate_gradient} explicitly.
		Since \eqref{ob:rate_passive} is a concave function of $\mathbf{\Theta}$, convergence to local-optimal points is guaranteed.
		On the other hand, the global optimal transmit precoder for a fixed $\mathbf{\Theta}$ is given by the eigenmode transmission \cite{Clerckx2013}
		\begin{equation}
			\mathbf{W}^\star = \mathbf{V} {\mathbf{S}^\star}^{1/2},
			\label{eq:precoder_eigenmode}
		\end{equation}
		where $\mathbf{V}$ is the right channel singular matrix and $\mathbf{S}^\star$ is the optimal water-filling power allocation matrix.
		The overall \gls{ao} algorithm converges to local-optimal points of problem \eqref{op:rate} since each subproblem is solved optimally and the objective is bounded from above.
	\end{subsection}

	\begin{subsection}{Low-Complexity Solution}
		We then propose a low-complexity solution to problem \eqref{op:rate} based on channel shaping.
		The passive beamforming subproblem \eqref{op:rate_passive} involves transmit covariance matrix $\mathbf{Q}$ and thus requires iterative \gls{rcg} update.
		Instead, we decouple the joint \gls{ris}-transceiver design by recasting \eqref{op:rate_passive} as channel power maximization
		\begin{maxi!}
			{\scriptstyle{\mathbf{\Theta}}}{\lVert \mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F} \rVert _\mathrm{F}^2}{\label{op:power_passive}}{\label{ob:power_passive}}
			\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g,}{}{}
		\end{maxi!}
		which only accounts for channel characteristics but has no trivial solution.
		Inspired by \cite{Nie2017}, we successively approximate the quadratic objective \eqref{ob:power_passive} by local Taylor expansions and solve each iteration in closed form.

		\begin{proposition}
			Start from any $\mathbf{\Theta}^{(0)} \in \mathbb{U}^{N_\mathrm{S} \times N_\mathrm{S}}$, the sequence
			\begin{equation}
				\mathbf{\Theta}_g^{(r+1)} = \mathbf{U}_g^{(r)} \mathbf{V}_g^{(r)}, \quad \forall g
			\end{equation}
			converges to a stationary point of \eqref{op:pc_power}, where $\mathbf{U}_g^{(r)}$ and $\mathbf{V}_g^{(r)}$ are the left and right compact singular matrix of
			\begin{equation}
				\mathbf{M}_g^{(r)} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} \Bigl(\mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathrm{diag}\bigl(\mathbf{\Theta}_{[1:g-1]}^{(r+1)},\mathbf{\Theta}_{[g:G]}^{(r)}\bigr) \mathbf{H}_\mathrm{F}\Bigr) \mathbf{H}_{\mathrm{F},g}^\mathsf{H}
			\end{equation}
		\end{proposition}

		\begin{proof}
			The differential of \eqref{ob:power_passive} w.r.t. $\mathbf{\Theta}_g^*$ is
			\begin{align*}
				\partial \lVert \mathbf{H} \rVert _\mathrm{F}^2
				& = \mathrm{tr}\bigl(\mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^* \cdot \mathbf{H}_{\mathrm{F},g}^* (\mathbf{H}_\mathrm{D}^\mathsf{T} + \mathbf{H}_\mathrm{F}^\mathsf{T} \mathbf{\Theta}^\mathsf{T} \mathbf{H}_\mathrm{B}^\mathsf{T})\bigr)\\
				& = \mathrm{tr}\bigl(\mathbf{H}_{\mathrm{F},g}^* (\mathbf{H}_\mathrm{D}^\mathsf{T} + \mathbf{H}_\mathrm{F}^\mathsf{T} \mathbf{\Theta}^\mathsf{T} \mathbf{H}_\mathrm{B}^\mathsf{T}) \mathbf{H}_{\mathrm{B},g}^* \cdot \partial \mathbf{\Theta}_g^*\bigr)
			\end{align*}
			and the corresponding complex derivative is
			\begin{equation}
				\frac{\partial \lVert \mathbf{H} \rVert _\mathrm{F}^2}{\partial \mathbf{\Theta}_g^*} = \mathbf{H}_{\mathrm{B},g}^\mathsf{H} (\mathbf{H}_\mathrm{D} + \mathbf{H}_\mathrm{B} \mathbf{\Theta} \mathbf{H}_\mathrm{F}) \mathbf{H}_{\mathrm{F},g}^\mathsf{H} = \mathbf{M}_g.
			\end{equation}
			% For a given $\mathbf{\Theta}$, we approximate the quadratic objective \eqref{ob:power_passive} by local Taylor expansion and recast problem \eqref{op:power_passive} as

			First, we approximate the quadratic objective \eqref{ob:power_passive} by its local Taylor expansion
			\begin{maxi!}
				{\scriptstyle{\mathbf{\Theta}}}{\sum_g 2 \Re\bigl\{ \mathrm{tr}(\mathbf{\Theta}_g^\mathsf{H} \mathbf{M}_g) \bigr\}}{\label{op:power_passive_approx}}{\label{ob:power_passive_approx}}
				\addConstraint{\mathbf{\Theta}_g^\mathsf{H} \mathbf{\Theta}_g=\mathbf{I}, \quad \forall g.}{}{}
			\end{maxi!}
			Let $\mathbf{M}_g = \mathbf{U}_g \mathbf{\Sigma}_g \mathbf{V}_g^\mathsf{H}$ be the compact \gls{svd} of $\mathbf{M}_g$.
			We have
			\begin{equation}
				\Re \bigl\{\mathrm{tr}(\mathbf{\Theta}_g^\mathsf{H} \mathbf{M}_g)\bigr\} = \Re \bigl\{ \mathrm{tr}(\mathbf{\Sigma}_g \mathbf{V}_g^\mathsf{H} \mathbf{\Theta}_g^\mathsf{H} \mathbf{U}_g) \bigr\} \le \mathrm{tr}(\mathbf{\Sigma}_g).
			\end{equation}
			The upper bound is tight when $\mathbf{V}_g^\mathsf{H} \mathbf{\Theta}_g^\mathsf{H} \mathbf{U}_g = \mathbf{I}$, which implies the optimal solution of \eqref{op:power_passive_approx} is $\tilde{\mathbf{\Theta}}_g = \mathbf{U}_g \mathbf{V}_g^\mathsf{H}$, $\forall g$.
